Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-28 paper/paper.tex
pass on section 3 (example)

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.
A Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.  In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. The header of this rule is \CodeIn{icmp any any -> any
  any}. It instructs Suricata to inspect \CodeIn{icmp} traffic flowing
from any machine and port to any machine and port.\MyComment{ The variables
\CodeIn{\$HOME\_NET} and \CodeIn{\$EXTERNAL\_NET} are
configurable.} The rule options come after the header. It is a
semi-colon-separated sequence of key-value pairs. Rule options serve
to document the analyzed traffic (\eg{}, ``msg'', ``classtype'', and
``sid'') and to characterize the attack pattern (\eg, ``flags'',
``threshold'', and ``content''). \tname{} produces options of the
second kind, which are \emph{relevant} to capture the attack. For
example, the option \CodeIn{content:<arg>} checks if the argument is
present in the payload of the message.

@@

This section briefly shows how \tname{} synthesizes rules for one
example attack. Active Reconnaissance is a method used by malicious
individuals to collect information about a computing system to
determine potential vulnerabilities. This method is used as a
preliminary step towards an actual attack. For example, Ping Scan
scans the network for IPs of accessible hosts. An attacker sends
several ICMP Echo requests to a range of IP addresses in the network
and checks which ones respond to these requests. This method enables
an attacker to know which machines are active in the network.

@@

Figure~\ref{fig:pingscan-example} shows a Suricata rule that captures
a Ping Scan attempt, as documented in NMAP~\cite{netmap}, a network
discovery and security auditing open-source tool. The relevant options
(see Section~\ref{sec:example-suricata-rules}) are in bold---the
option \CodeIn{dsize: 0} indicates that the packet has no bytes in the
payload and the option \CodeIn{itype: 8} indicates that the packet to
be captured are ICMP Echo requests.

In the following, we briefly show how \tname{} synthesizes rules for
this attack. The action and the header of the rules that \tname{}
creates are either inferred or left constant. For example, in this
case, \tname{} sets the action to \CodeIn{alert} and sets the header
to \CodeIn{<proto> any any -> any any}, indicating that it infers the
target protocol and uses the most general flow description possible,
\ie{}, it instructs the \nids\ to analyze the traffic flowing from/to
any address and port.  The rationale for this decision is that the
choice of these parameters are domain-specific and subject to change
by the system administrator.  The main challenge in synthesizing rules
is to discover the set of options in a rule that blocks the malicious
traffic, but allows the benign traffic to pass. \Luc{O Suricata normalmente 
nao bloqueia trafego, apenas gera alertas sobre ele. Para bloquear um 
pacote ele precisa ser configurado no modo Intrusion Prevention System 
(IPS) ao inves de IDS: https://blog.starti.com.br/ids-ips/}

\tname{} proceeds as follows to synthesize rules. First, it analyzes
the malicious traffic and extracts options expressed in that
traffic. Recall from Section~\ref{sec:rules-and-packets} that some
options can be inferred from packet contents. \tname{} finds the
following options at this stage:
{\scriptsize{\texttt{\textbf{dsize}:0; \textbf{itype:}8;
      \textbf{icode:}0; \textbf{icmp\_id:}23570;
      \textbf{icmp\_seq:}3439;}}}.  By construction, a rule that uses
this set of options captures the malicious traffic. Unfortunately, the
rule above is overspecified. Informally, the rule includes ``too
specific'' characteristics of the malicious traffic and could
potentially result in false negatives. In this case, although the
options \CodeIn{dsize:0} and \CodeIn{itype:8} are present in the set
of inferred options (as the golden rule from
Figure~\ref{fig:pingscan-example}), the others are not.

@@

Second, \tname{} uses the benign traffic to mitigate the overfitting
issue. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive malicious traffic but
not capture the negative benign traffic. It uses the rule obtained in
the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options from the seed rule. Considering this example, \tname{}
produces a total of \pingscanplausible{} plausible rules at the end of
the search.  In the worst case, \tname{} produces $2^n$ plausible
rules, where $n$ is the number of options contained in the inferred
rule ($n=5$, in this case). Rules are discared in the search process
as they are found to be implausible.

@@

Finally, \tname{} ranks plausible rules according to heuristics
learned from a database of existing rules. The intuition is that new
rules should share characteristics with pre-existing rules. For
example, one heuristic function used by \tname{} leverages the
\emph{size of a rule} for that. The score of a rule given by such a
function is proportional to the popularity of the number of options
declared in rules of a given
protocol. Figure~\ref{fig:distribution-rule-size} shows the
distribution of the number of content options associated with rules
from a public data set~\cite{emerging-threats-open}. Note that rules
with 2 to 6 options are more common. At the end of this step, \tname{}
ranks the golden rule at the top of the ranking, including
\pingscanplausible{} rules. It is worth noting that
\tname\ consistently ranks the golden rule very high in most of the
cases even when the seed rule, produced at the first step, contains a
large number of options (see Table~\ref{table:results}).

>>>>

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.
A Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.  In this example, a
message will be sent to system administrators if the rule pattern is
satisfied\footnote{Technically, an IPS (Intrusion Prevention System)
  differs from an IDS in the action taken when an intrusion is
  observed. IPS systems block malicous traffic. However, IDS and IPS
  do not differ in the approaches that can be used to detect the
  intrusion~\cite{ids-ips}.\Mar{check}}. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. The header of this rule is \CodeIn{icmp any any -> any
  any}. It instructs Suricata to inspect \CodeIn{icmp} traffic flowing
from any machine and port to any machine and port.\MyComment{ The variables
\CodeIn{\$HOME\_NET} and \CodeIn{\$EXTERNAL\_NET} are
configurable.} The rule options come after the header. It is a
semi-colon-separated sequence of key-value pairs. Rule options serve
to document the analyzed traffic (\eg{}, ``msg'', ``classtype'', and
``sid'') and to characterize the attack pattern (\eg, ``flags'',
``threshold'', and ``content''). \tname{} produces options of the
second kind, which are \emph{relevant} to capture the attack. For
example, the option \CodeIn{content:<arg>} checks if the argument is
present in the payload of the message.

@@

This section briefly shows how \tname{} synthesizes rules for one
example attack. Active Reconnaissance is typically used as a
preliminary step towards an actual attack. It is a method to collect
information about a computing system to determine potential
vulnerabilities. For example, Ping Scan scans the network for IPs of
accessible hosts. An attacker sends several ICMP Echo requests to a
range of IP addresses in the network and checks which ones respond to
these requests. This method enables an attacker to discover which
machines are active in the network.

@@

Figure~\ref{fig:pingscan-example} shows a Suricata rule that captures
a Ping Scan, as documented in NMAP~\cite{netmap}, a network discovery
and security auditing open-source tool. The essential options to
capture the attack (see Section~\ref{sec:example-suricata-rules}) are
in bold---the option \CodeIn{dsize: 0} indicates that the packet has
no bytes in the payload and the option \CodeIn{itype: 8} indicates
that the packet to be captured are ICMP Echo requests.

In the following, we briefly show how \tname{} synthesizes rules for
this attack. The action is left constant (\CodeIn{alert}) and the
header is inferred from the protocol. For example, in this case,
\tname{} sets the header to \CodeIn{icmp any any -> any any},
indicating that the rule applies to \CodeIn{icmp} messages and applies
to the most general flow description, \ie{}, it instructs the
\nids\ to analyze the traffic flowing from/to any address and port.
The rationale for this decision is that the choice of these parameters
are domain-specific and subject to change by the system administrator.

The main challenge in synthesizing rules is to discover the set of
rule options that capture the malicious traffic, but allow the benign
traffic to pass. \tname{} proceeds as follows to synthesize rule
options. First, it analyzes the malicious traffic to extract options
expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that some options can be inferred
from packet contents. \tname{} finds the following five options at
this stage: {\scriptsize{\texttt{\textbf{dsize}:0; \textbf{itype:}8;
      \textbf{icode:}0; \textbf{icmp\_id:}23570;
      \textbf{icmp\_seq:}3439;}}}.  By construction, a rule that uses
this set of options captures the malicious traffic. Unfortunately, the
rule above is overspecified. Informally, the rule includes ``too
specific'' characteristics of the malicious traffic and could
potentially result in false negatives. In this case, although the
options \CodeIn{dsize:0} and \CodeIn{itype:8} are present in the set
of inferred options (as the golden rule from
Figure~\ref{fig:pingscan-example}), the others are not.

@@

Second, \tname{} uses the benign traffic to mitigate the overfitting
problem. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive malicious traffic but
not capture the negative benign traffic. It uses the rule obtained in
the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options declared in the seed rule. Considering this example,
\tname{} produces a total of \pingscanplausible{}\Fix{check!}
plausible rules at the end of the search.  Note that there are $2^n$
possible plausible rules, where $n$ is the number of options contained
in the seed rule ($n=5$, in this case). \tname{} is able to prune the
search space and discard many implausible rules. It is worth noting
that the number of plausible rules that \tname\ produces at this step
can be very high for other example attacks (see column ``\#~Plausible
Rules'' on Table~\ref{table:results}).

@@

Finally, \tname{} ranks plausible rules according to heuristics
obtained by analyzing a database of existing rules. The general
intuition is that new rules should share characteristics with
pre-existing rules. For example, one heuristic function used by
\tname{} leverages the \emph{size of a rule} for ranking.  The score
of a rule $r$ that this function produces is proportional to the
popularity of the number of options in
$r$. Figure~\ref{fig:distribution-rule-size} shows the distribution of
the number of content options associated with rules from a public data
set~\cite{emerging-threats-open}. Note that rules with 2 to 6 options
are more common. At the end of this step, \tname{} ranks the golden
rule at the top of the ranking, including \pingscanplausible{}
rules. It is worth noting that \tname\ consistently ranks the golden
rule very high in most of the cases even when the technique produces a
large number of plausible rules (see Table~\ref{table:results}).
